Maximum Entropy Spectral Analysis: a case study

### 2023-07-11 paper/mesa.tex
added place holder text and figures

  \abstract{
The Maximum Entropy Spectral Analysis (MESA) method, developed by Burg, provides a powerful tool to perform spectral estimation of a time-series. The method relies on a Jaynes' maximum entropy principle and provides the means of inferring the spectrum of a stochastic process in terms of the coefficients of some autoregressive process AR($p$) of order $p$.
A closed form recursive solution provides an estimate of the autoregressive coefficients as well as of the order $p$ of the process.
We provide a ready-to-use implementation of the algorithm in the form of a python package \texttt{memspectrum}. We characterize our implementation by performing a power spectral density analysis on synthetic data (with known power spectral density) and we compare different criteria for stopping the recursion. Furthermore, we compare the performance of our code with the ubiquitous Welch algorithm, using synthetic data generated from the released spectrum by the LIGO-Virgo collaboration.
We find that, when compared to Welch's method, Burg's method provides a power spectral density (PSD) estimation with a 
systematically lower variance and bias. This is particularly manifest in the case of a little number of data points, making Burg's method most suitable to work in this regime.}

@@

Our paper is organized as follows: we begin by briefly reviewing the theoretical foundations of the maximum entropy principle in Sec.~\ref{sec:foundations}. Sec.~\ref{sec:validation} presents the validation of Burg's method as well as of our implementation on simulated data. In Sec.~\ref{sec:Welch_comparison} we compare the results from \texttt{memspectrum} with the Welch method; Sec.~\ref{sec:applications} presents a few applications to real time series and, finally, we conclude with a discussion in Sec.~\ref{sec:conclusion}.

@@

\section{Applications} \label{sec:applications}
\subsection{Temperature Time Series} \label{sec:temp_timeseries}
As a further example of the breadth of applicability of MESA, we applied our implementation to atmospheric temperature time series. 
The reason for this choice is twofold: i) atmospheric temperature time series present a variety of overlapping periodicities most of which are 
known; ii) it provides a stress test for the time series forecast analysis.  As dataset, we  used the historical reanalysis data 
from the ``ERA5-Land hourly data from 1981 to present" dataset, downloaded from the Climate Data Store~\citep{ERA5_dataset}.
The data consist in temperatures taken at coordinates $\textrm{N}\;\ang{45;5;} \;\; \textrm{E}\;\ang{9;1;}}$, corresponding about to the city of Milan, Italy. The temperatures are given on an hourly cadence for almost 31 years from 31st December 1989 to 30th November 2020.

Fig.~\ref{fig:temp_spectrum}, shows the MESA spectrum inferred from the data. As a comparison, we adopt both the FPE and the CAT loss functions.
In agreement with what we found in previous sections, the FPE PSD is more regular compared to the one from CAT. Both spectra show a
peak at the frequency $f_D = \SI{1}{day^{-1}}$ corresponding to one day, corresponding to the day-night cycle: this is expected.
Higher order harmonics (corresponding to integer multiples of $f_D$) are also visible up to the Nyquist frequency ($f_D = \SI{0.5}{hour^{-1}}$): they correspond to signals that preserve the 1 day period while, at the same time, capturing the complex variability of the daily temperature cycle throughout the year.

Looking at low frequencies, FPE does not capture the yearly variability at $f_{yr} = \SI{1}{yr^{-1}}$. On the other hand, the peak is captured well by the CAT loss function. In analogy as what observed for the peak at $f_D$, in the CAT spectrum we observe a higher order harmonic at a frequency $2 \cdot f_{yr}$. As in the previous case, this is required to better model the temperature variation in the year.

The failure of FPE in capturing the low frequency trend can be understood by looking at Fig.~\ref{fig:p_vs_ptrue}. It is shown that CAT systematically select very long filters (i.e. large $p_{CAT}$ values for the autoregressive orders), whereas FPE tends to select smaller values for the order $p_{FPE}$ of the autoregressive process, often underestimating the actual value.
In this example, we have $p_{CAT} \sim  36500 \sim  \SI{4}{yr}$, whereas $p_{FPE} \sim  1300 \sim  \SI{2}{months}$.
The autoregressive process selected by FPE, hence cannot produce a meaningful prediction on the timescale longer than a few weeks and for this reason it is unable to capture the low frequency peak at $f_{yr}$. In other words, the model is too simple to model \textit{both} the behaviour at high and low frequencies (separated by approximately 3 orders of magnitude).
On the other hand, the model chosen by CAT is much more complex and is able to model also the high frequency behaviour, at the expense of making a more noisy estimation.
The ``actual" length of the autoregressive process should be closer to the choice of CAT.

We now assess the accuracy of the forecasting of new observations of the time series. 
Based on actual data, we try to predict future values as described in sec.~\ref{sec:forecasting} (of course the MESA is performed with CAT loss function). We produce $N =100$ independent predictions and we compute the median as well as the $90\%$ confidence interval. This is compared with the actual measured temperature values. The predictions span a two years range of time. We report our results in Fig.~\ref{fig:temp_spectrum}.

We note the observed difference is always well included in the $90\%$ confidence interval: the forecasting predictions seem reliable.
On the other hand, the confidence interval is pretty large (almost as large as $\SI{15}{K}$), thus making ``easy" for the actual data to fit the predictions.
Indeed, the prediction model, while suitable for spectral estimation, is nothing more than a linear regression (plus noise term). Such a simple model hardly catches the complex trend in the variability of the temperature daily trend during a year. For more precise predictions, probably one should consider nonlinear regression model, tapping into the wealth of non linear predictors offered by the field of Deep Learning.

\begin{figure}
	\caption{Spectrum for the historical temperature time series. The two lines refers to different loss functions (CAT and FPE); the inset shows the harmonics of the fundamental frequency of a day. Two vertical lines are drawn in correspondence to the frequency $f_D$ of a day and $f_{yr}$ of a year.}
	\label{fig:temp_spectrum}
	\includegraphics{Images/climate_plots/temp_spectrum.pdf}
\end{figure}

\begin{figure}
	\caption{Difference between the historical temperature $T$ and the forecast temperature $T_{forecast}$ for two years of data. The red shadow denotes the $90\%$ confidence interval of the predictions. The model is trained with CAT loss function. The discrepancy between predicted and actual values is always everywhere included in the $90\%$ confidence interval.}
	\label{fig:temp_forecast}
	\includegraphics{Images/climate_plots/forecast_accuracy.pdf}
\end{figure}

>>>>

  \abstract{
The Maximum Entropy Spectral Analysis (MESA) method, developed by Burg, provides a powerful tool to perform spectral estimation of a time-series. The method relies on a Jaynes' maximum entropy principle and provides the means of inferring the spectrum of a stochastic process in terms of the coefficients of some autoregressive process AR($p$) of order $p$.
A closed form recursive solution provides an estimate of the autoregressive coefficients as well as of the order $p$ of the process.
We provide a ready-to-use implementation of the algorithm in the form of a python package \texttt{memspectrum}. We characterize our implementation by performing a power spectral density analysis on synthetic data (with known power spectral density) and we compare different criteria for stopping the recursion. Furthermore, we compare the performance of our code with the ubiquitous Welch algorithm, using synthetic data generated from the released spectrum by the LIGO-Virgo-Kagra collaboration.
We find that, when compared to Welch's method, Burg's method provides a power spectral density (PSD) estimation with a 
systematically lower variance and bias. This is particularly manifest in the case of a little number of data points, making Burg's method most suitable to work in this regime. 
We conclude our investigation by utilising MESA, and its particularly easy parametrisation where the only free parameter is the order $p$ of the AR process, to marginalise over the interferometers noise PSD in conjunction with inferring the parameters of GW150914.}

@@

Our paper is organized as follows: we begin by briefly reviewing the theoretical foundations of the maximum entropy principle in Sec.~\ref{sec:foundations}. Sec.~\ref{sec:validation} presents the validation of Burg's method as well as of our implementation on simulated data. In Sec.~\ref{sec:Welch_comparison} we compare the results from \texttt{memspectrum} with the Welch method; Sec.~\ref{sec:applications} presents a few applications to real time series, including the analysis of GW150914, and, finally, we conclude with a discussion in Sec.~\ref{sec:conclusion}.

@@

\section{Applications} \label{sec:applications}
%\subsection{Temperature Time Series} \label{sec:temp_timeseries}
%As a further example of the breadth of applicability of MESA, we applied our implementation to atmospheric temperature time series. 
%The reason for this choice is twofold: i) atmospheric temperature time series present a variety of overlapping periodicities most of which are 
%known; ii) it provides a stress test for the time series forecast analysis.  As dataset, we  used the historical reanalysis data 
%from the ``ERA5-Land hourly data from 1981 to present" dataset, downloaded from the Climate Data Store~\citep{ERA5_dataset}.
%The data consist in temperatures taken at coordinates $\textrm{N}\;\ang{45;5;} \;\; \textrm{E}\;\ang{9;1;}}$, corresponding about to the city of Milan, Italy. The temperatures are given on an hourly cadence for almost 31 years from 31st December 1989 to 30th November 2020.
%
%Fig.~\ref{fig:temp_spectrum}, shows the MESA spectrum inferred from the data. As a comparison, we adopt both the FPE and the CAT loss functions.
%In agreement with what we found in previous sections, the FPE PSD is more regular compared to the one from CAT. Both spectra show a
%peak at the frequency $f_D = \SI{1}{day^{-1}}$ corresponding to one day, corresponding to the day-night cycle: this is expected.
%Higher order harmonics (corresponding to integer multiples of $f_D$) are also visible up to the Nyquist frequency ($f_D = \SI{0.5}{hour^{-1}}$): they correspond to signals that preserve the 1 day period while, at the same time, capturing the complex variability of the daily temperature cycle throughout the year.
%
%Looking at low frequencies, FPE does not capture the yearly variability at $f_{yr} = \SI{1}{yr^{-1}}$. On the other hand, the peak is captured well by the CAT loss function. In analogy as what observed for the peak at $f_D$, in the CAT spectrum we observe a higher order harmonic at a frequency $2 \cdot f_{yr}$. As in the previous case, this is required to better model the temperature variation in the year.
%
%The failure of FPE in capturing the low frequency trend can be understood by looking at Fig.~\ref{fig:p_vs_ptrue}. It is shown that CAT systematically select very long filters (i.e. large $p_{CAT}$ values for the autoregressive orders), whereas FPE tends to select smaller values for the order $p_{FPE}$ of the autoregressive process, often underestimating the actual value.
%In this example, we have $p_{CAT} \sim  36500 \sim  \SI{4}{yr}$, whereas $p_{FPE} \sim  1300 \sim  \SI{2}{months}$.
%The autoregressive process selected by FPE, hence cannot produce a meaningful prediction on the timescale longer than a few weeks and for this reason it is unable to capture the low frequency peak at $f_{yr}$. In other words, the model is too simple to model \textit{both} the behaviour at high and low frequencies (separated by approximately 3 orders of magnitude).
%On the other hand, the model chosen by CAT is much more complex and is able to model also the high frequency behaviour, at the expense of making a more noisy estimation.
%The ``actual" length of the autoregressive process should be closer to the choice of CAT.
%
%We now assess the accuracy of the forecasting of new observations of the time series. 
%Based on actual data, we try to predict future values as described in sec.~\ref{sec:forecasting} (of course the MESA is performed with CAT loss function). We produce $N =100$ independent predictions and we compute the median as well as the $90\%$ confidence interval. This is compared with the actual measured temperature values. The predictions span a two years range of time. We report our results in Fig.~\ref{fig:temp_spectrum}.
%
%We note the observed difference is always well included in the $90\%$ confidence interval: the forecasting predictions seem reliable.
%On the other hand, the confidence interval is pretty large (almost as large as $\SI{15}{K}$), thus making ``easy" for the actual data to fit the predictions.
%Indeed, the prediction model, while suitable for spectral estimation, is nothing more than a linear regression (plus noise term). Such a simple model hardly catches the complex trend in the variability of the temperature daily trend during a year. For more precise predictions, probably one should consider nonlinear regression model, tapping into the wealth of non linear predictors offered by the field of Deep Learning.
%
%\begin{figure}
%	\caption{Spectrum for the historical temperature time series. The two lines refers to different loss functions (CAT and FPE); the inset shows the harmonics of the fundamental frequency of a day. Two vertical lines are drawn in correspondence to the frequency $f_D$ of a day and $f_{yr}$ of a year.}
%	\label{fig:temp_spectrum}
%	\includegraphics{Images/climate_plots/temp_spectrum.pdf}
%\end{figure}
%
%\begin{figure}
%	\caption{Difference between the historical temperature $T$ and the forecast temperature $T_{forecast}$ for two years of data. The red shadow denotes the $90\%$ confidence interval of the predictions. The model is trained with CAT loss function. The discrepancy between predicted and actual values is always everywhere included in the $90\%$ confidence interval.}
%	\label{fig:temp_forecast}
%	\includegraphics{Images/climate_plots/forecast_accuracy.pdf}
%\end{figure}

\subsection{Marginalising over the noise PSD in GW parameter estimation: the case for GW150914}
As an application of MESA, and its implementation in \texttt{memspectrum}, we analyse GW150914\cite{} using a Bayesian framework that allows for the marginalisation of the order  $p$ of the AR($p$) process representing the noise data stream. The inference algorithm proceeds as usual, see \cite{}[veitch and vecchio, ashton et al, etc etc], with a modification to the likelihood calculation. The likelihood function we employ is given in Eq.~\eqref{eq:Whittle} and its calculation proceeds as follows: for all proposed Monte Carlo moves $\theta\equiv\theta_{GW}\cup p$\footnote{We indicate the set of all GW parameters (component masses, spins, luminosity distance, etc.) with $\theta_{GW}$.}, for each detector included in the analysis, we compute the detector response to the template $h(t;\theta_{GW})$ in the time-domain, subtract it from the data $d(t)$ to get the time domain residuals $r(t;\theta_{GW})$. Under the MESA assumption, we describe the PSD of the residuals as in Eq.~\eqref{eq:MESApsd}, where the order $p$ is fixed by the Monte Carlo sampler. Having computed the PSD, we Fourier transform the residuals and compute the likelihood. Our sampler is a nested sampling algorithm\cite{}[raynest] and the specific inference model is implemented as part of \texttt{granite}\cite{}[in preparation], a dedicated inference model for ground based interferometric detectors. Needless to say, despite the implementation of the fast Burg algorithm in \texttt{memspectrum}, the inference with noise marginalisation is vastly more computational expensive than the standard analysis. As our analysis template, we adopt the machine learning based MLGW model\cite{}[MLGW], an aligned spin model trained on TEOBResumS\cite{}, that has been shown to perform well on LVK detected events\cite{}[MLGW]. 
\begin{figure}
	\caption{In the top panel, we show the difference between the forecasted strain values (CAT model) and the actual values observed at Livingston interferometer (blue). The red shadow denotes the $90\%$ confidence interval subtracted with the actual time series values. The confidence interval is computed on  $100$ independent random realization of the predictions.
	The bottom panel displays the standard deviation of the predictions with both CAT and FPE models. The inset shows a detail of the first $\SI{0.5}{s}$ of plot in the top panel.
	}
	\label{fig:gw150914_intrinsic}
	\includegraphics{Images/gw150914/intrinsic.pdf}
\end{figure}
\begin{figure}
	\caption{In the top panel, we show the difference between the forecasted strain values (CAT model) and the actual values observed at Livingston interferometer (blue). The red shadow denotes the $90\%$ confidence interval subtracted with the actual time series values. The confidence interval is computed on  $100$ independent random realization of the predictions.
	The bottom panel displays the standard deviation of the predictions with both CAT and FPE models. The inset shows a detail of the first $\SI{0.5}{s}$ of plot in the top panel.
	}
	\label{fig:gw150914_intrinsic}
	\includegraphics{Images/gw150914/extrinsic.pdf}
\end{figure}
\begin{figure}
	\caption{In the top panel, we show the difference between the forecasted strain values (CAT model) and the actual values observed at Livingston interferometer (blue). The red shadow denotes the $90\%$ confidence interval subtracted with the actual time series values. The confidence interval is computed on  $100$ independent random realization of the predictions.
	The bottom panel displays the standard deviation of the predictions with both CAT and FPE models. The inset shows a detail of the first $\SI{0.5}{s}$ of plot in the top panel.
	}
	\label{fig:gw150914_intrinsic}
	\includegraphics{Images/gw150914/waveforms.pdf}
\end{figure}
\begin{figure}
	\caption{In the top panel, we show the difference between the forecasted strain values (CAT model) and the actual values observed at Livingston interferometer (blue). The red shadow denotes the $90\%$ confidence interval subtracted with the actual time series values. The confidence interval is computed on  $100$ independent random realization of the predictions.
	The bottom panel displays the standard deviation of the predictions with both CAT and FPE models. The inset shows a detail of the first $\SI{0.5}{s}$ of plot in the top panel.
	}
	\label{fig:gw150914_intrinsic}
	\includegraphics{Images/gw150914/posterior_order_H1.pdf}
\end{figure}

\begin{figure}
	\caption{In the top panel, we show the difference between the forecasted strain values (CAT model) and the actual values observed at Livingston interferometer (blue). The red shadow denotes the $90\%$ confidence interval subtracted with the actual time series values. The confidence interval is computed on  $100$ independent random realization of the predictions.
	The bottom panel displays the standard deviation of the predictions with both CAT and FPE models. The inset shows a detail of the first $\SI{0.5}{s}$ of plot in the top panel.
	}
	\label{fig:gw150914_intrinsic}
	\includegraphics{Images/gw150914/posterior_order_L1.pdf}
\end{figure}

\begin{figure}
	\caption{In the top panel, we show the difference between the forecasted strain values (CAT model) and the actual values observed at Livingston interferometer (blue). The red shadow denotes the $90\%$ confidence interval subtracted with the actual time series values. The confidence interval is computed on  $100$ independent random realization of the predictions.
	The bottom panel displays the standard deviation of the predictions with both CAT and FPE models. The inset shows a detail of the first $\SI{0.5}{s}$ of plot in the top panel.
	}
	\label{fig:gw150914_intrinsic}
	\includegraphics{Images/gw150914/noise_curves.pdf}
\end{figure}

In Figs.~\ref{fig:gw150914_intrinsic}, ~\ref{fig:gw150914_extrinsic} and \ref{fig:gw150914_wf} we show the posteriors for the set of intrinsic parameters, extrinsic parameters and reconstructed waveform from our analysis. In addition, Fig.~ \ref{fig:gw150914_orders} shows the posterior distribution on the orders $p_{H}$ and $p_{L}$ for the AR processes describing the residuals in the LIGO-Hanford and LIGO-Livingston detectors. The inferred PSDs are instead shown in Fig.~\ref{fig:gw150914_psds}.
